---
title: "Tendon Biomechanics"
subtitle: "Clustering"
author: "Emily Johnson"
date: today
date-format: short
format:
  html:
    self-contained: true
    theme: litera
    toc: true
editor: source
code-block-bg: true
code-block-border-left: "#31BAE9"
---

# Failure profile clustering

Failure profiles are classified using DTW clustering on the full failure curve (absolute force in N, absolute displacement in mm), interpolated onto a common grid spanning the observed displacement range. This preserves both curve shape and scale.

Sample inclusion is taken from the results summary written by `01_biomechanical_processing.py`: a sample enters clustering only if it passes the QC column selected below. QC is therefore defined once, in the processing script, rather than recomputed inline here.

## Load libraries

```{r}
#| output: false
#| warning: false
#| message: false
library(tidyverse)
library(dtwclust)
library(signal)
library(cowplot)
library(doParallel)
library(doRNG)

filter <- dplyr::filter  # signal::filter masks dplyr::filter
```

## Parameters

Tunable parameters and colour palette. SG window and poly match the processing script's failure-region settings (window = 101, poly = 3).

```{r}
params <- list(
  # Preprocessing
  onset_fraction   = 0.05,   # force threshold as fraction of peak-to-trough range
  onset_points     = 5L,     # consecutive points required to confirm onset
  sg_window        = 101L,    # Savitzky-Golay window 
  sg_poly          = 3L,     # Savitzky-Golay poly

  # DTW clustering
  n_grid           = 300L,   # interpolation grid points
  dtw_window       = 0.10,   # Sakoe-Chiba band as fraction of n_grid
  k_range          = 2:7,    # cluster counts to evaluate

  # Feature-based clustering
  post_peak_n      = 20L     # points used to estimate post-peak drop rate
)


bright <- c(
  "#EE7733",
  "#BB2222",
  "#0077BB",
  "#66CCEE",
  "#009988",
  "#AA4499",
  "#CCBB44"   
)
```

## Load data

It should be noted that `compiled_Force_N_fail_only.csv` and `compiled_Displacement_mm_fail_only.csv` were produced by filtering the main data files to the failure segments. The reason for this was that the files were too large/unwiedly to easily read into R.

Readers can apply the same prefiltering by running:

`grep -i "fail" compiled_Displacement_mm_phase_140526.1.csv >> compiled_Displacement_mm_fail_only.csv`

`grep -i "fail" compiled_Force_N_phase_140526.1.csv >> compiled_Force_N_fail_only.csv`

```{r}

meta_raw    <- read.csv("raw_data/compiled_metadata_with_filenames.csv")
results_raw <- read.csv("results/results_summary_050826.csv")
force_raw   <- read.csv("raw_data/compiled_Force_N_fail_only.csv")
disp_raw    <- read.csv("raw_data/compiled_Displacement_mm_fail_only.csv")

```

## Process metadata

```{r}

meta <- meta_raw |>
  rename(MouseID = 3) |>   # mouse ID column is literally named "." in the CSV
  mutate(
    sample_key = str_remove(FileName, "\\.csv$"),
    sample_key = gsub(" ", ".", sample_key)
  ) |>
  distinct(sample_key, .keep_all = TRUE)

glimpse(meta)

```

## Failure QC from the processing script

`SNR failure pass` is the SNR filter applied to the failure segment in `01_biomechanical_processing.py`. Swap `qc_column` to `Failure.pass` for the combined SNR / residual / range gate; the exclusion labels below follow whichever is set.

```{r}

qc_column <- "SNR.failure.pass"

qc <- results_raw |>
  transmute(
    sample_key   = gsub(" ", ".", str_remove(File.name, "\\.csv$")),
    failure_pass = as.logical(.data[[qc_column]])
  ) |>
  distinct(sample_key, .keep_all = TRUE)

qc_pass_keys <- qc |> filter(failure_pass) |> pull(sample_key)

cat("Samples in results summary:", nrow(qc), "\n")
cat(qc_column, "= TRUE:", length(qc_pass_keys), "\n")

```

## Extract per-sample vectors

Each compiled file contains triplets of columns per sample (SetName, Cycle, value). SetName and value columns are matched by suffix regex. Non-failure rows are masked to NA; the result is a named list of numeric vectors. Note: this is because we are only comparing the shape of failure profiles here, we could do full profiles over time as an expanded analysis...

```{r}

extract_sample_vectors <- function(raw_df, value_suffix_regex) {
  cols        <- names(raw_df)
  setnam_cols <- cols[str_detect(cols, regex("_SetNam?$", ignore_case = TRUE))]
  value_cols  <- cols[str_detect(cols, regex(value_suffix_regex, ignore_case = TRUE))]

  stopifnot(length(setnam_cols) == length(value_cols))
  cat("Samples detected:", length(setnam_cols), "\n")

  map2(setnam_cols, value_cols, \(sc, vc) {
    is_failure <- str_detect(as.character(raw_df[[sc]]),
                             regex("fail", ignore_case = TRUE))
    vals       <- suppressWarnings(as.numeric(raw_df[[vc]]))
    vals[!is_failure] <- NA
    vals
  }) |>
    set_names(
      setnam_cols |>
        str_remove("^X") |>
        str_remove(regex("_SetNam?$", ignore_case = TRUE))
    )
}

force_vecs <- extract_sample_vectors(force_raw, "_Force_?$")
disp_vecs  <- extract_sample_vectors(disp_raw,  "_Displacement_?$")

# Restrict to samples present in both DFs and accompanying metadata, then to
# those passing failure QC in the results summary
common_keys <- Reduce(intersect, list(names(force_vecs),
                                      names(disp_vecs),
                                      meta$sample_key))

qc_keys <- intersect(common_keys, qc_pass_keys)

cat("Samples with paired traces and metadata:", length(common_keys), "\n")
cat("Excluded by failure QC:", length(setdiff(common_keys, qc_keys)), "\n")
cat("Samples entering preprocessing:", length(qc_keys), "\n")

```

## Preprocessing

Steps matched to the processing script:

1.  Retain finite paired observations
2.  Baseline correction: subtract initial force value
3.  Smooth with Savitzky-Golay (window 101, poly 3)
4.  Onset detection on smoothed trace
5.  Trim from onset to end of recording to maintain post-peak shape

Displacement and force are retained in absolute units (mm and N). Signal-quality exclusion comes from the QC column selected above. The remaining `NULL` returns are structural e.g a curve too short to smooth, or with no detectable onset, cannot be interpolated onto the grid at all.

```{r}

find_onset <- function(f_smooth, threshold, n_consec) {
  runs      <- rle(f_smooth > threshold)
  run_end   <- cumsum(runs$lengths)
  run_start <- c(1L, run_end[-length(run_end)] + 1L)
  long_runs <- which(runs$values & runs$lengths >= n_consec)
  if (length(long_runs) == 0L) return(NA_integer_)
  run_start[long_runs[1]]
}

preprocess_curve <- function(force_vec, disp_vec) {
  # Retain finite pairs
  keep <- is.finite(force_vec) & is.finite(disp_vec)
  f    <- force_vec[keep]
  d    <- disp_vec[keep]
  if (length(f) < params$sg_window + 50L) return(NULL)

  # Baseline correction
  f <- f - f[1]
  d <- d - d[1]

  # Smooth
  f_smooth <- sgolayfilt(f, p = params$sg_poly, n = params$sg_window)

  # Onset detection
  peak_idx   <- which.max(f_smooth)
  if (is.na(peak_idx) || peak_idx < params$sg_window) return(NULL)

  pre_peak   <- f_smooth[seq_len(peak_idx)]
  trough     <- min(pre_peak)
  rise_range <- max(pre_peak) - trough
  if (rise_range <= 0) return(NULL)

  threshold <- trough + params$onset_fraction * rise_range
  onset_idx <- find_onset(pre_peak, threshold, params$onset_points)
  if (is.na(onset_idx) || onset_idx >= peak_idx) return(NULL)

  # Trim onset to end of recording (post-peak retained)
  f_trim <- f_smooth[onset_idx:length(f_smooth)]
  d_trim <- d[onset_idx:length(d)]
  # d_trim <- d_trim - d_trim[1]  # uncomment to align curves at onset 
  peak_i <- peak_idx - onset_idx + 1L

  if (length(f_trim) < params$sg_window || peak_i < 3L) return(NULL)

  list(
    f_series = f_trim,
    d_series = d_trim,
    peak_i   = peak_i
  )
}

processed <- map(
  qc_keys,
  \(k) preprocess_curve(force_vecs[[k]], disp_vecs[[k]])
) |>
  set_names(qc_keys)


n_dropped <- sum(map_lgl(processed, is.null))
cat("Series dropped after preprocessing:", n_dropped, "\n")
processed <- keep(processed, ~ !is.null(.x))
cat("Series entering clustering:", length(processed), "\n")

```

------------------------------------------------------------------------

# DTW clustering

Force (N) and displacement (mm) are in absolute units. Each curve is interpolated onto a common displacement grid spanning the 95th percentile of observed failure displacement (this avoids a small number of outlier samples stretching the grid and wasting resolution on padding). Samples with displacement shorter than the grid ceiling are padded by repeating their final observed force value.

```{r}
all_max_disp <- map_dbl(processed, \(p) max(p$d_series, na.rm = TRUE))
grid_max     <- quantile(all_max_disp, 0.95)
d_grid       <- seq(0, grid_max, length.out = params$n_grid)

interpolate_to_grid <- function(proc) {
  approx(proc$d_series, proc$f_series,
         xout = d_grid,
         rule = 2,
         ties = "ordered")$y
}

series_dtw <- map(processed, interpolate_to_grid)

cat("Grid max displacement:", round(grid_max, 2), "mm\n")
cat("All series length", params$n_grid, ":",
    all(map_int(series_dtw, length) == params$n_grid), "\n")
```

## Fit

```{r}
n_cores  <- detectCores() - 1L
cl       <- makeCluster(n_cores)
registerDoParallel(cl)
clusterEvalQ(cl, library(dtwclust))

set.seed(42)
registerDoRNG(42)   # foreach workers need their own RNG streams
win_size <- as.integer(params$n_grid * params$dtw_window)

dtw_fits <- map(params$k_range, \(k) {
  t0  <- proc.time()
  fit <- tsclust(
    series   = series_dtw,
    type     = "partitional",
    k        = k,
    distance = "dtw_basic",
    centroid = "pam",
    trace    = FALSE,
    args     = tsclust_args(dist = list(window.size = win_size))
  )
  cat("k =", k, "done in",
      round((proc.time() - t0)["elapsed"] / 60, 1), "min\n")
  fit
}) |>
  set_names(paste0("k", params$k_range))

stopCluster(cl)
registerDoSEQ()
```

## Cluster validation

CH, DB, and D are computed across all k. Silhouette is computed only for the consensus best k, given the cost of pairwise DTW distance computation at n = `r length(series_dtw)`.

```{r}
cvi_df_dtw <- map_dfr(params$k_range, \(k) {
  fit <- dtw_fits[[paste0("k", k)]]
  tibble(
    k  = k,
    DB = cvi(fit, type = "DB"),
    CH = cvi(fit, type = "CH"),
    D  = cvi(fit, type = "D")
  )
})

print(cvi_df_dtw)

best_k_dtw <- cvi_df_dtw |>
  mutate(
    rank_DB   = rank(DB),
    rank_CH   = rank(-CH),
    rank_D    = rank(-D),
    mean_rank = (rank_DB + rank_CH + rank_D) / 3
  ) |>
  slice_min(mean_rank, n = 1, with_ties = FALSE) |>
  pull(k)

# best_k_dtw <- 6L  # override with biological prior if needed
cat("Best k — DTW (consensus):", best_k_dtw, "\n")

dtw_best <- dtw_fits[[paste0("k", best_k_dtw)]]
sil_dtw  <- cvi(dtw_best, type = "Sil")
cat("Silhouette at best k:", round(sil_dtw, 3), "\n")

```

## Cluster assignments

```{r}

assign_dtw <- tibble(
  sample_key  = names(series_dtw),
  cluster_dtw = as.integer(dtw_best@cluster)
) |>
  left_join(meta, by = "sample_key")

processed_keys <- names(processed)

# cluster_dtw = 0 for samples available but excluded, 100 for samples with no
# usable failure trace; exclusion_reason records which gate dropped them
excluded_df <- bind_rows(
  tibble(sample_key       = setdiff(common_keys, qc_keys),
         exclusion_reason = paste(qc_column, "not TRUE")),
  tibble(sample_key       = setdiff(qc_keys, processed_keys),
         exclusion_reason = "preprocessing failed")
) |>
  mutate(cluster_dtw = 0) |>
  left_join(meta, by = "sample_key")

not_processed_df <- tibble(
  sample_key       = setdiff(meta$sample_key, common_keys),
  cluster_dtw      = 100,
  exclusion_reason = "no paired failure trace"
) |>
  left_join(meta, by = "sample_key")

assign_dtw$cluster_dtw <- as.numeric(assign_dtw$cluster_dtw)

assign_dtw_full <- bind_rows(
  assign_dtw,
  excluded_df,
  not_processed_df
)

count(assign_dtw_full, exclusion_reason)


dir.create("figures", showWarnings = FALSE)

write.csv(
  assign_dtw_full,
  file = file.path("results", "dtw_cluster_assignments.csv"),
  row.names = FALSE
)

assign_dtw |>
  count(cluster_dtw, Genotype, Sex, Age) |>
  arrange(cluster_dtw)

assign_dtw <- assign_dtw |>
  mutate(
    Genotype = factor(
      Genotype,
      levels = c(
        "oim Wt", "oim Het", "oim Hom",
        "Tm1b Wt", "Tm1b Het", "Tm1b Hom"
      ),
      labels = c(
        "J +/+", "J +/oim", "J oim/oim",
        "N +/+", "N +/-", "N -/-"
      )
    )
)


```

## Plots

### Validation indices by k

```{r fig.width=12, fig.height=4}

ch_p <- ggplot(cvi_df_dtw, aes(x = k, y = CH)) +
  geom_line(colour = "#31BAE9") +
  geom_point(size = 3, colour = "#31BAE9") +
  geom_vline(xintercept = best_k_dtw, linetype = "dashed", colour = "grey50") +
  labs(
    title = paste0("DTW clustering: CH index by k  (best k = ", best_k_dtw,
                   ", Sil = ", round(sil_dtw, 3), ")"),
    x = "k", y = "Calinski-Harabasz index"
  ) +
  theme_cowplot()

ch_p
save_plot("figures/dtw_cvi.pdf", ch_p, base_width = 12, base_height = 4)

```

### Centroids with individual curves

```{r fig.width=9, fig.height=3}

# QC-fail samples never entered clustering, so preprocess and interpolate them
# here to show what is being excluded (structural failures return NULL)
series_fail <- map(setdiff(common_keys, qc_keys),
                   \(k) preprocess_curve(force_vecs[[k]], disp_vecs[[k]])) |>
  set_names(setdiff(common_keys, qc_keys)) |>
  keep(~ !is.null(.x)) |>
  map(interpolate_to_grid)

cat("QC-fail samples shown:", length(series_fail),
    "of", length(setdiff(common_keys, qc_keys)), "\n")

cluster_ids    <- sort(unique(assign_dtw$cluster_dtw))
assign_colours <- c(set_names(bright[seq_along(cluster_ids)], as.character(cluster_ids)),
                    "QC fail" = "grey55")

centroid_df_dtw <- imap_dfr(dtw_best@centroids, \(cen, idx) {
  tibble(
    disp_mm = d_grid,
    force_n = as.numeric(cen),
    cluster = as.character(idx)
  )
})

member_df_dtw <- bind_rows(
  imap_dfr(series_dtw, \(s, key) {
    tibble(disp_mm = d_grid, force_n = as.numeric(s), sample_key = key)
  }) |>
    left_join(assign_dtw |> select(sample_key, cluster_dtw), by = "sample_key") |>
    mutate(cluster = as.character(cluster_dtw)),
  imap_dfr(series_fail, \(s, key) {
    tibble(disp_mm = d_grid, force_n = as.numeric(s), sample_key = key,
           cluster = "QC fail")
  })
) |>
  mutate(cluster = factor(cluster, levels = names(assign_colours)))

cluster_labels <- assign_dtw |>
  count(cluster_dtw) |>
  mutate(
    cluster_dtw = as.character(cluster_dtw),
    label       = paste0("Cluster ", cluster_dtw,
                         "\n(n = ", n, ")")
  ) |>
  select(cluster_dtw, label) |>
  deframe() |>
  c("QC fail" = paste0("QC fail\n(n = ", length(series_fail), ")"))

centroid_p_dtw <- ggplot(mapping = aes(x = disp_mm, y = force_n, colour = cluster)) +
  geom_line(
    data      = member_df_dtw,
    aes(group = sample_key),
    alpha     = 0.05,
    linewidth = 0.15
  ) +
  geom_line(data = centroid_df_dtw, linewidth = 0.7) +
  facet_wrap(
    ~ cluster, nrow = 1, scales = "fixed",
    labeller = labeller(cluster = cluster_labels)
  ) +
  scale_colour_manual(values = assign_colours) +
  scale_y_continuous(
    limits = c(0, max(member_df_dtw$force_n) * 1.05),
    expand = expansion(mult = c(0, 0))
  ) +
  labs(x = "Displacement (mm)", y = "Force (N)") +
  theme_cowplot() +
  theme(
    legend.position    = "none",
    strip.background   = element_blank(),
    strip.text         = element_text(size = 12),
    panel.grid.major.y = element_line(colour = "grey92", linewidth = 0.3),
    panel.spacing      = unit(0.8, "lines")
  )

centroid_p_dtw
save_plot("figures/dtw_centroids.pdf", centroid_p_dtw, base_width = 9, base_height = 3)

```

### Cluster composition by genotype

Composition is taken from `assign_dtw_full`, so samples excluded by QC appear alongside the clusters. Samples with no compiled failure trace are dropped entirely, so denominators are all samples of that genotype that were actually measured, making the QC-fail fraction directly comparable across groups.

```{r}

comp_df <- assign_dtw_full |>
  dplyr::filter(cluster_dtw != 100) |>
  mutate(
    Assignment = if_else(cluster_dtw == 0, "QC fail", as.character(cluster_dtw)),
    Assignment = factor(Assignment, levels = names(assign_colours)),
    Genotype = factor(
      Genotype,
      levels = c(
        "oim Wt", "oim Het", "oim Hom",
        "Tm1b Wt", "Tm1b Het", "Tm1b Hom"
      ),
      labels = c(
        "J +/+", "J +/oim", "J oim/oim",
        "N +/+", "N +/-", "N -/-"
      )
    ),
    Age = factor(Age, levels = c("8wks", "18wks", "52wks"))
  )

comp_df |>
  count(Genotype, Assignment) |>
  pivot_wider(names_from = Assignment, values_from = n, values_fill = 0)

```

```{r fig.width=7, fig.height=3.5}

comp_dtw <- comp_df |>
  count(Assignment, Genotype) |>
  ggplot(aes(y = Genotype, x = n, fill = Assignment)) +
  geom_col(position = "fill", width = 0.7) +
  scale_x_continuous(labels = scales::percent_format(),
                     expand = expansion(mult = c(0, 0.02))) +
  scale_fill_manual(values = assign_colours) +
  labs(x = "Proportion", y = NULL, fill = "Assignment") +
  theme_cowplot() +
  theme(
    legend.position = "right",
    axis.line.y     = element_blank(),
    axis.ticks.y    = element_blank()
  )

comp_dtw
save_plot("figures/dtw_composition.pdf", comp_dtw, base_width = 7, base_height = 3.5)

```

```{r fig.width=9, fig.height=4}

comp_dtw_facet <- comp_df |>
  dplyr::filter(!is.na(Age)) |>
  count(Assignment, Genotype, Age) |>
  ggplot(aes(x = Age, y = n, fill = Assignment)) +
  geom_col(position = "fill", width = 0.55) +
  scale_y_continuous(
    labels = scales::percent_format(),
    expand = expansion(mult = c(0, 0.02))
  ) +
  scale_fill_manual(values = assign_colours) +
  facet_wrap(~ Genotype, nrow = 2, ncol = 3) +
  labs(x = NULL, y = "Proportion", fill = "Assignment") +
  theme_cowplot() +
  theme(
    strip.background   = element_blank(),
    strip.text         = element_text(size = 11),
    panel.grid.major.y = element_line(colour = "grey92", linewidth = 0.3),
    panel.spacing      = unit(0.8, "lines"),
    axis.text.x        = element_text(size = 9.5)
  )

comp_dtw_facet
save_plot("figures/dtw_composition_by_age.pdf", comp_dtw_facet,
          base_width = 9, base_height = 4)

```

::: {.callout-note collapse="true"}
## Session Info

```{r}
sessionInfo()
```
:::
