---
title: "Tendon Biomechanics"
subtitle: "Clustering"
author: "Emily Johnson"
date: today
date-format: short
format:
  html:
    self-contained: true
    theme: litera
    toc: true
editor: source
code-block-bg: true
code-block-border-left: "#31BAE9"
---

# Failure profile clustering

Failure profiles are classified using DTW clustering on the full failure curve (absolute force in N, absolute displacement in mm), interpolated onto a common grid spanning the observed displacement range. This preserves both curve shape and scale.

## Load libraries

```{r}
#| output: false
#| warning: false
#| message: false
library(tidyverse)
library(readxl)
library(dtwclust)
library(signal)
library(cowplot)
library(ggrepel)
library(doParallel)
library(factoextra)
library(cluster)

filter <- dplyr::filter  # signal::filter masks dplyr::filter
```

## Parameters

Tunable parameters and colour palette. SG window and poly match the processing script's failure-region settings (window = 101, poly = 3).

```{r}
params <- list(
  # Preprocessing
  onset_fraction   = 0.05,   # force threshold as fraction of peak-to-trough range
  onset_points     = 5L,     # consecutive points required to confirm onset
  sg_window        = 101L,    # Savitzky-Golay window 
  sg_poly          = 3L,     # Savitzky-Golay poly
  min_snr          = 3.0,    # min peak/noise ratio; excludes distorted profiles

  # DTW clustering
  n_grid           = 300L,   # interpolation grid points
  dtw_window       = 0.10,   # Sakoe-Chiba band as fraction of n_grid
  k_range          = 2:7,    # cluster counts to evaluate

  # Feature-based clustering
  post_peak_n      = 20L     # points used to estimate post-peak drop rate
)


bright <- c(
  "#EE7733",
  "#BB2222",
  "#0077BB",
  "#66CCEE",
  "#009988",
  "#AA4499",
  )
```

## Load data

```{r}
meta_raw  <- read.csv("compiled_metadata_with_filenames.csv")
force_raw <- read.csv("compiled_Force_N_fail_only.csv")
disp_raw  <- read.csv("compiled_Displacement_mm_fail_only.csv")
```

## Process metadata

```{r}
meta <- meta_raw |>
  mutate(
    sample_key = str_remove(FileName, "\\.csv$"),
    sample_key = gsub(" ", ".", sample_key)
  ) |>
  distinct(sample_key, .keep_all = TRUE)

glimpse(meta)
```

## Extract per-sample vectors

Each compiled file contains triplets of columns per sample (SetName, Cycle, value). SetName and value columns are matched by suffix regex. Non-failure rows are masked to NA; the result is a named list of numeric vectors.

```{r}

extract_sample_vectors <- function(raw_df, value_suffix_regex) {
  cols        <- names(raw_df)
  setnam_cols <- cols[str_detect(cols, regex("_SetNam?$", ignore_case = TRUE))]
  value_cols  <- cols[str_detect(cols, regex(value_suffix_regex, ignore_case = TRUE))]

  stopifnot(length(setnam_cols) == length(value_cols))
  cat("Samples detected:", length(setnam_cols), "\n")

  map2(setnam_cols, value_cols, \(sc, vc) {
    is_failure <- str_detect(as.character(raw_df[[sc]]),
                             regex("fail", ignore_case = TRUE))
    vals       <- suppressWarnings(as.numeric(raw_df[[vc]]))
    vals[!is_failure] <- NA
    vals
  }) |>
    set_names(
      setnam_cols |>
        str_remove("^X") |>
        str_remove(regex("_SetNam?$", ignore_case = TRUE))
    )
}

force_vecs <- extract_sample_vectors(force_raw, "_Force_?$")
disp_vecs  <- extract_sample_vectors(disp_raw,  "_Displacement_?$")

# Restrict to samples present in both DFs and accompanying metadata before processing
common_keys <- Reduce(intersect, list(names(force_vecs),
                                      names(disp_vecs),
                                      meta$sample_key))

cat("Samples entering preprocessing:", length(common_keys), "\n")

```

## Preprocessing

Steps matched to the processing script:

1. Retain finite paired observations
2. Baseline correction: subtract initial force value
3. Smooth with Savitzky-Golay (window 101, poly 3)
4. Onset detection on smoothed trace
5. Trim from onset to end of recording to maintain post-peak shape
6. SNR filter: exclude distorted/low-force profiles

Displacement and force are retained in absolute units (mm and N).

```{r}
find_onset <- function(f_smooth, threshold, n_consec) {
  runs      <- rle(f_smooth > threshold)
  run_end   <- cumsum(runs$lengths)
  run_start <- c(1L, run_end[-length(run_end)] + 1L)
  long_runs <- which(runs$values & runs$lengths >= n_consec)
  if (length(long_runs) == 0L) return(NA_integer_)
  run_start[long_runs[1]]
}

preprocess_curve <- function(force_vec, disp_vec) {
  # Retain finite pairs
  keep <- is.finite(force_vec) & is.finite(disp_vec)
  f    <- force_vec[keep]
  d    <- disp_vec[keep]
  if (length(f) < params$sg_window + 50L) return(NULL)

  # Baseline correction
  f <- f - f[1]
  d <- d - d[1]

  # Smooth
  f_smooth <- sgolayfilt(f, p = params$sg_poly, n = params$sg_window)

  # Onset detection
  peak_idx   <- which.max(f_smooth)
  if (is.na(peak_idx) || peak_idx < params$sg_window) return(NULL)

  pre_peak   <- f_smooth[seq_len(peak_idx)]
  trough     <- min(pre_peak)
  rise_range <- max(pre_peak) - trough
  if (rise_range <= 0) return(NULL)

  threshold <- trough + params$onset_fraction * rise_range
  onset_idx <- find_onset(pre_peak, threshold, params$onset_points)
  if (is.na(onset_idx) || onset_idx >= peak_idx) return(NULL)

  # Trim onset to end of recording (post-peak retained)
  f_trim <- f_smooth[onset_idx:length(f_smooth)]
  d_trim <- d[onset_idx:length(d)]
  peak_i <- peak_idx - onset_idx + 1L

  if (length(f_trim) < params$sg_window || peak_i < 3L) return(NULL)

  # SNR filter
  noise_sd <- sd(f[seq_len(min(50L, onset_idx - 1L))])
  if (!is.na(noise_sd) && noise_sd > 0 &&
      (f_trim[peak_i] / noise_sd) < params$min_snr) return(NULL)

  list(
    f_series = f_trim,
    d_series = d_trim,
    peak_i   = peak_i
  )
}

processed <- map(
  common_keys,
  \(k) preprocess_curve(force_vecs[[k]], disp_vecs[[k]])
) |>
  set_names(common_keys)


n_dropped <- sum(map_lgl(processed, is.null))
cat("Series dropped after preprocessing:", n_dropped, "\n")
processed <- keep(processed, ~ !is.null(.x))
cat("Series entering clustering:", length(processed), "\n")

```

---

# DTW clustering

Force (N) and displacement (mm) are in absolute units. Each curve is interpolated onto a common displacement grid spanning the 95th percentile of observed failure displacement — this avoids a small number of outlier specimens stretching the grid and wasting resolution on padding. Specimens with displacement shorter than the grid ceiling are padded by repeating their final observed force value.

```{r}
all_max_disp <- map_dbl(processed, \(p) max(p$d_series, na.rm = TRUE))
grid_max     <- quantile(all_max_disp, 0.95)
d_grid       <- seq(0, grid_max, length.out = params$n_grid)

interpolate_to_grid <- function(proc) {
  approx(proc$d_series, proc$f_series,
         xout = d_grid,
         rule = 2,
         ties = "ordered")$y
}

series_dtw <- map(processed, interpolate_to_grid)

cat("Grid max displacement:", round(grid_max, 2), "mm\n")
cat("All series length", params$n_grid, ":",
    all(map_int(series_dtw, length) == params$n_grid), "\n")
```

## Fit

```{r}
n_cores  <- detectCores() - 1L
cl       <- makeCluster(n_cores)
registerDoParallel(cl)
clusterEvalQ(cl, library(dtwclust))

set.seed(42)
win_size <- as.integer(params$n_grid * params$dtw_window)

dtw_fits <- map(params$k_range, \(k) {
  t0  <- proc.time()
  fit <- tsclust(
    series   = series_dtw,
    type     = "partitional",
    k        = k,
    distance = "dtw_basic",
    centroid = "pam",
    trace    = FALSE,
    args     = tsclust_args(dist = list(window.size = win_size))
  )
  cat("k =", k, "done in",
      round((proc.time() - t0)["elapsed"] / 60, 1), "min\n")
  fit
}) |>
  set_names(paste0("k", params$k_range))

stopCluster(cl)
registerDoSEQ()
```

## Cluster validation

CH, DB, and D are computed across all k. Silhouette is computed only for the consensus best k, given the cost of pairwise DTW distance computation at n = `r length(series_dtw)`.

```{r}
cvi_df_dtw <- map_dfr(params$k_range, \(k) {
  fit <- dtw_fits[[paste0("k", k)]]
  tibble(
    k  = k,
    DB = cvi(fit, type = "DB"),
    CH = cvi(fit, type = "CH"),
    D  = cvi(fit, type = "D")
  )
})

print(cvi_df_dtw)

best_k_dtw <- cvi_df_dtw |>
  mutate(
    rank_DB   = rank(DB),
    rank_CH   = rank(-CH),
    rank_D    = rank(-D),
    mean_rank = (rank_DB + rank_CH + rank_D) / 3
  ) |>
  slice_min(mean_rank, n = 1) |>
  pull(k)

# best_k_dtw <- 6L  # override with biological prior if needed
cat("Best k — DTW (consensus):", best_k_dtw, "\n")

dtw_best <- dtw_fits[[paste0("k", best_k_dtw)]]
sil_dtw  <- cvi(dtw_best, type = "Sil")
cat("Silhouette at best k:", round(sil_dtw, 3), "\n")
```

## Cluster assignments

```{r}
assign_dtw <- tibble(
  sample_key  = names(series_dtw),
  cluster_dtw = as.integer(dtw_best@cluster)
) |>
  left_join(meta, by = "sample_key")

processed_keys <- names(processed)
removed_keys   <- setdiff(common_keys, processed_keys)

filtered_df <- tibble(
  sample_key  = removed_keys,
  cluster_dtw = 0
) |>
  left_join(meta, by = "sample_key")

not_processed_keys <- setdiff(meta$sample_key, common_keys)

not_processed_df <- tibble(
  sample_key  = not_processed_keys,
  cluster_dtw = 100
) |>
  left_join(meta, by = "sample_key")

assign_dtw$cluster_dtw <- as.numeric(assign_dtw$cluster_dtw)

assign_dtw_full <- bind_rows(
  assign_dtw,
  filtered_df,
  not_processed_df
)

names(assign_dtw_full)[5] <- "MouseID"

dir.create("outputs", showWarnings = FALSE)

write.csv(
  assign_dtw_full,
  file = file.path("outputs", "dtw_cluster_assignments.csv"),
  row.names = FALSE
)

assign_dtw |>
  count(cluster_dtw, Genotype, Sex, Age) |>
  arrange(cluster_dtw)

genotype_levels <- c("oim Hom", "oim Het", "oim Wt", "Tm1b Hom", "Tm1b Het", "Tm1b Wt")

assign_dtw <- assign_dtw |>
  mutate(
    Genotype = factor(
      Genotype,
      levels = c(
        "oim Wt", "oim Het", "oim Hom",
        "Tm1b Wt", "Tm1b Het", "Tm1b Hom"
      ),
      labels = c(
        "J +/+", "J +/oim", "J oim/oim",
        "N +/+", "N +/-", "N -/-"
      )
    )
)


```

## Plots

### Validation indices by k

```{r fig.width=7, fig.height=4}
ch_p <- ggplot(cvi_df_dtw, aes(x = k, y = CH)) +
  geom_line(colour = "#31BAE9") +
  geom_point(size = 3, colour = "#31BAE9") +
  geom_vline(xintercept = best_k_dtw, linetype = "dashed", colour = "grey50") +
  labs(
    title = paste0("DTW clustering: CH index by k  (best k = ", best_k_dtw,
                   ", Sil = ", round(sil_dtw, 3), ")"),
    x = "k", y = "Calinski-Harabasz index"
  ) +
  theme_cowplot()

ch_p
save_plot("figures/dtw_cvi.pdf", ch_p, base_width = 12, base_height = 4)
```

### Centroids with individual curves

```{r fig.width=11, fig.height=5}

centroid_df_dtw <- imap_dfr(dtw_best@centroids, \(cen, idx) {
  tibble(
    disp_mm = d_grid,
    force_n = as.numeric(cen),
    cluster = as.character(idx)
  )
})

member_df_dtw <- imap_dfr(series_dtw, \(s, key) {
  clust <- assign_dtw |> filter(sample_key == key) |> pull(cluster_dtw)
  tibble(
    disp_mm    = d_grid,
    force_n    = as.numeric(s),
    cluster    = as.character(clust),
    sample_key = key
  )
})

cluster_labels <- assign_dtw |>
  count(cluster_dtw) |>
  mutate(
    cluster_dtw = as.character(cluster_dtw),
    label       = paste0("Cluster ", cluster_dtw,
                         "\n(n = ", n, ")")
  ) |>
  select(cluster_dtw, label) |>
  deframe()
centroid_p_dtw <- ggplot(mapping = aes(x = disp_mm, y = force_n, colour = cluster)) +
  geom_line(
    data      = member_df_dtw,
    aes(group = sample_key),
    alpha     = 0.03,
    linewidth = 0.15
  ) +
  geom_line(data = centroid_df_dtw, linewidth = 0.7) +
  facet_wrap(
    ~ cluster, nrow = 1, scales = "fixed",
    labeller = labeller(cluster = cluster_labels)
  ) +
  scale_colour_manual(values = bright) +
  scale_y_continuous(
    limits = c(0, max(member_df_dtw$force_n) * 1.05),
    expand = expansion(mult = c(0, 0))
  ) +
  labs(x = "Displacement (mm)", y = "Force (N)") +
  theme_cowplot() +
  theme(
    legend.position    = "none",
    strip.background   = element_blank(),
    strip.text         = element_text(size = 12),
    panel.grid.major.y = element_line(colour = "grey92", linewidth = 0.3),
    panel.spacing      = unit(0.8, "lines")
  )

centroid_p_dtw
save_plot("figures/dtw_centroids.pdf", centroid_p_dtw, base_width = 7.5, base_height = 3)

```

### Cluster composition by genotype

```{r fig.width=7, fig.height=5}

comp_dtw <- assign_dtw |>
  count(cluster_dtw, Genotype) |>
  mutate(cluster_dtw = factor(cluster_dtw)) |>
  ggplot(aes(y = Genotype, x = n, fill = cluster_dtw)) +
  geom_col(position = "fill", width = 0.7) +
  scale_x_continuous(labels = scales::percent_format(),
                     expand = expansion(mult = c(0, 0.02))) +
  scale_fill_manual(values = bright) +
  labs(x = "Proportion", y = NULL, fill = "Cluster") +
  theme_cowplot() +
  theme(
    legend.position = "right",
    axis.line.y     = element_blank(),
    axis.ticks.y    = element_blank()
  )

comp_dtw
save_plot("figures/dtw_composition.pdf", comp_dtw, base_width = 12, base_height = 3.5)

```

```{r}

comp_dtw_facet <- assign_dtw |>
  count(cluster_dtw, Genotype, Age) |>
  mutate(
    cluster_dtw = factor(cluster_dtw),
    Age         = factor(Age, levels = c("8wks", "18wks", "52wks"))
  ) |>
  dplyr::filter(!is.na(Age)) |>
  ggplot(aes(x = Age, y = n, fill = cluster_dtw)) +
  geom_col(position = "fill", width = 0.65) +
  scale_y_continuous(
    labels = scales::percent_format(),
    expand = expansion(mult = c(0, 0.02))
  ) +
  scale_fill_manual(values = bright) +
  facet_wrap(~ Genotype, nrow = 2, ncol = 3) +
  labs(x = NULL, y = "Proportion", fill = "Cluster") +
  theme_cowplot() +
  theme(
    strip.background   = element_blank(),
    strip.text         = element_text(size = 11),
    panel.grid.major.y = element_line(colour = "grey92", linewidth = 0.3),
    panel.spacing      = unit(0.8, "lines"),
    axis.text.x        = element_text(size = 9.5)
  )

comp_dtw_facet
save_plot("figures/dtw_composition_by_age.pdf", comp_dtw_facet,
          base_width = 12, base_height = 4.5)

```

::: {.callout-note collapse="true"}
## Session Info
```{r}
sessionInfo()
```
:::
