---
title: "Tendon Biomechanics"
subtitle: "Summary plots for paper"
author: "Emily Johnson"
date: today
date-format: short
format:
  html:
    self-contained: true
    theme: litera
    toc: true
editor: source
code-block-bg: true
code-block-border-left: "#31BAE9"
---

# Summary plots

## Libraries

```{r}
#| output: false
#| warning: false
#| message: false
library(tidyverse)
library(ggbeeswarm)   # beeswarm / quasirandom points
library(cowplot)      # theme_cowplot(), save_plot()
library(ggtext)       # italic gene names in strips
library(glue)
```

## Assay definitions

One CSV per assay per timepoint. Headers contain en-dashes, so columns are matched by pattern and renamed. `metrics` maps the renamed column to its plot label; `figure` gives the output file stem.

```{r}
data_dir   <- "./processed_data"
timepoints <- c("8wks", "18wks", "52wks")

assays <- list(
  diameter = list(
    prefix = "Tendon-diameter", stamp = "260626.5",
    rename = c("Diameter (um)" = "Average diameter (um)"),
    metrics = c("Average diameter (um)" = "Diameter (um)"),
    figure  = c("Diameter (um)" = "diameter")
  ),
  hysteresis = list(
    prefix = "Hysteresis", stamp = "080726.2",
    rename = c(
      "Hysteresis strain energy density (MPa)" = "Hysteresis strain energy per unit volume",
      "Hysteresis (%)"                         = "Hysteresis % (cycle 1 classical"
    ),
    metrics = c(
      "Hysteresis strain energy per unit volume" = "Hysteresis strain energy density (MPa)",
      "Hysteresis % (cycle 1 classical"          = "Hysteresis (%)"
    ),
    figure = c(
      "Hysteresis strain energy density (MPa)" = "hysteresis_MPa",
      "Hysteresis (%)"                         = "hysteresis_percent"
    )
  ),
  stress_relax = list(
    prefix = "Stress-relax", stamp = "080726.2",
    rename = c(
      "Rate of change of stress (MPa/s)" = "Rate of change of stress 60s",
      "Stress relaxation (%)"            = "Stress-relaxation 60s"
    ),
    metrics = c(
      "Rate of change of stress 60s" = "Rate of change of stress (MPa/s)",
      "Stress-relaxation 60s"        = "Stress relaxation (%)"
    ),
    figure = c(
      "Rate of change of stress (MPa/s)" = "rate_change_stress",
      "Stress relaxation (%)"            = "stress_relax_percent"
    )
  ),
  failure = list(
    prefix = "Failure", stamp = "080726.2",
    rename = c(
      "Failure stress"          = "failure_stress",
      "Failure strain"          = "failure_strain",
      "Max modulus"             = "max_modulus",
      "Failure force"           = "failure_force",
      "Failure extension"       = "failure_extension",
      "Stress at max modulus"   = "stress_max_modulus",
      "Strain at max modulus"   = "strain_max_modulus"
    ),
    metrics = c(
      failure_stress      = "Failure stress (MPa)",
      failure_strain      = "Failure strain (%)",
      max_modulus         = "Maximum modulus (MPa)",
      failure_force       = "Failure force (N)",
      failure_extension   = "Failure extension (mm)",
      stress_max_modulus  = "Stress at maximum modulus (MPa)",
      strain_max_modulus  = "Strain at maximum modulus (%)"
    ),
    figure = c(
      "Failure stress (MPa)"            = "failure_stress",
      "Failure strain (%)"              = "failure_strain",
      "Maximum modulus (MPa)"           = "max_modulus",
      "Failure force (N)"               = "failure_force",
      "Failure extension (mm)"          = "failure_extension",
      "Stress at maximum modulus (MPa)" = "stress_at_maxmod",
      "Strain at maximum modulus (%)"   = "strain_at_maxmod"
    )
  )
)
```

## Read and tidy

Each genotype carries two pieces of information (line and zygosity); these are recombined into allele-precise x-axis labels.

```{r}
#| warning: false
#| message: false

# First matching pattern wins; unmatched names are left alone.
rename_cols <- function(nms, map) {
  map <- c(map, c("Mouse ID" = "mouse_id"))
  map_chr(nms, \(nm) {
    hit <- which(str_detect(nm, names(map)))[1]
    if (is.na(hit)) nm else unname(map[hit])
  })
}

add_metadata <- function(df) {
  df |>
    mutate(
      Line = factor(if_else(str_starts(Genotype, "oim"), "oim", "Col1a2-null"),
                    levels = c("oim", "Col1a2-null")),
      Zygosity = case_when(
        str_detect(Genotype, "Wt")  ~ "Wt",
        str_detect(Genotype, "Het") ~ "Het",
        str_detect(Genotype, "Hom") ~ "Hom"
      ),
      Genotype = factor(
        paste(Line, Zygosity),
        levels = c("oim Wt", "oim Het", "oim Hom",
                   "Col1a2-null Wt", "Col1a2-null Het", "Col1a2-null Hom")
      ),
      Allele = factor(case_when(
        Line == "oim"         & Zygosity == "Wt"  ~ "+/+",
        Line == "oim"         & Zygosity == "Het" ~ "+/oim",
        Line == "oim"         & Zygosity == "Hom" ~ "oim/oim",
        Line == "Col1a2-null" & Zygosity == "Wt"  ~ "+/+",
        Line == "Col1a2-null" & Zygosity == "Het" ~ "+/−",
        Line == "Col1a2-null" & Zygosity == "Hom" ~ "−/−"
      ), levels = c("+/+", "+/oim", "oim/oim", "+/−", "−/−")),
      # "M J +/+" etc: sex, line (J = oim, N = Col1a2-null), allele.
      # Note: The sex prefix only orders/separates the groups and is stripped from the final tick labels.
      AlleleLabel = factor(
        paste(Sex, if_else(Line == "oim", "J", "N"), Allele),
        levels = c(
          "M J +/+", "M J +/oim", "M J oim/oim",
          "F J +/+", "F J +/oim", "F J oim/oim",
          "M N +/+", "M N +/−", "M N −/−",
          "F N +/+", "F N +/−", "F N −/−"
        )
      ),
      Sex = factor(Sex, levels = c("F", "M"), labels = c("Female", "Male")),
      Timepoint = factor(Age, levels = timepoints,
                         labels = c("8 weeks", "18 weeks", "52 weeks"))
    )
}

read_assay <- function(spec) {
  set_names(timepoints) |>
    map(\(tp) {
      read_csv(file.path(data_dir, glue("{spec$prefix}_{tp}_{spec$stamp}.csv")),
               locale = locale(encoding = "latin1"),
               show_col_types = FALSE) |>
        rename_with(rename_cols, map = spec$rename)
    }) |>
    list_rbind() |>
    add_metadata() |>
    select(mouse_id, Sex, Line, Genotype, Allele, AlleleLabel, Timepoint,
           all_of(names(spec$metrics))) |>
    pivot_longer(all_of(names(spec$metrics)),
                 names_to = "Metric", values_to = "value") |>
    filter(!is.na(value)) |>
    mutate(Metric = unname(spec$metrics[Metric]))
}

metric_order <- unlist(map(assays, \(s) unname(s$metrics)), use.names = FALSE)
figure_stems <- do.call(c, unname(map(assays, "figure")))

dat_long <- assays |>
  map(read_assay) |>
  list_rbind(names_to = "Assay") |>
  mutate(Metric = factor(Metric, levels = metric_order))

```

Each mouse has several technical replicates but only ~4 mice per group, and the replicates aren't independent, so summaries use per-mouse means. Raw replicates appear only as a faint background layer. In the statistical analysis mixed models were employed to account for pseudo-replication, treating each mouse as a biological unit.

```{r}

# One point per mouse 
mouse_means <- dat_long |>
  summarise(value = mean(value),
            .by = c(mouse_id, Sex, Line, Genotype, Allele, AlleleLabel,
                    Timepoint, Assay, Metric))

# Mice per group
mouse_means |>
  distinct(Timepoint, Genotype, Sex, mouse_id) |>
  count(Timepoint, Genotype, Sex) |>
  pivot_wider(names_from = Sex, values_from = n, values_fill = 0)

# Mean and SD, with whiskers clipped at zero for strictly non-negative metrics
group_stats <- mouse_means |>
  summarise(
    mean = mean(value),
    sd = sd(value),
    min_value = min(value),
    max_value = max(value),
    .by = c(AlleleLabel, Timepoint, Metric)
  ) |>
  mutate(
    line_lower = if_else(min_value >= 0, pmax(mean - sd, 0), mean - sd),
    line_upper = if_else(max_value <= 0, pmin(mean + sd, 0), mean + sd)
  )
```

## Plotting

Points are coloured per mouse within each group (warm palette for females, cool for males), so replicates can be traced back to their animal.

```{r}

female_cols <- c(
  "#E69F00",
  "#D55E00",
  "#CC79A7",
  "#C2185B"
)

male_cols <- c(
  "#0072B2",  
  "#56B4E9",  
  "#3949AB",  
  "#6A3D9A"   
)

mouse_cols <- mouse_means |>
  distinct(mouse_id, Sex, Timepoint, Genotype) |>
  arrange(Timepoint, Genotype, Sex, mouse_id) |>
  group_by(Timepoint, Genotype, Sex) |>
  mutate(mouse_col = if (first(Sex) == "Female") {
    female_cols[seq_len(n())]
  } else {
    male_cols[seq_len(n())]
  }) |>
  ungroup()

join_by_mouse <- c("mouse_id", "Sex", "Timepoint", "Genotype")
mouse_means <- left_join(mouse_means, mouse_cols, by = join_by_mouse)
dat_long    <- left_join(dat_long,    mouse_cols, by = join_by_mouse)

```

One helper draws a metric across all three timepoints. `style` picks the point layer: `"tech"` adds the faint technical replicates behind the per-mouse means (default); `"beeswarm"` and `"jitter"` show the means only.

```{r}

plot_metric <- function(metric_name, style = c("tech", "beeswarm", "jitter")) {

  style <- match.arg(style)
  md <- filter(mouse_means, Metric == metric_name)
  td <- filter(dat_long,    Metric == metric_name)
  gd <- filter(group_stats, Metric == metric_name)

  p <- ggplot(md, aes(AlleleLabel, value))

  if (style == "tech") {
    p <- p + geom_quasirandom(data = td, aes(colour = mouse_col),
                              width = 0.15, size = 1, alpha = 0.35, stroke = 0)
  }

  p <- p + switch(style,
    jitter = geom_jitter(aes(colour = mouse_col), width = 0.1, height = 0,
                         size = 2.5, stroke = 0),
    geom_beeswarm(aes(colour = mouse_col), size = 2.5, cex = 2.2, stroke = 0)
  )

  # mean +/- SD: vertical range line, wide bar at the mean, narrow bars at ± SD
  crossbar <- function(y, width) {
    geom_crossbar(data = gd, aes(x = AlleleLabel, y = {{ y }},
                                 ymin = {{ y }}, ymax = {{ y }}),
                  inherit.aes = FALSE, width = width,
                  colour = "grey35", linewidth = 0.2)
  }

  p +
    geom_segment(data = gd,
                 aes(x = AlleleLabel, xend = AlleleLabel,
                     y = line_lower, yend = line_upper),
                 inherit.aes = FALSE, colour = "grey35", linewidth = 0.2) +
    crossbar(mean, 0.6) +
    crossbar(mean + sd, 0.4) +
    crossbar(mean - sd, 0.4) +
    facet_grid(. ~ Timepoint, scales = "free_x", space = "free_x") +
    scale_colour_identity() +
    scale_x_discrete(drop = TRUE, labels = \(x) gsub("^[MF] ", "", x)) +
    scale_y_continuous(limits = c(min(0, min(td$value, na.rm = TRUE)), NA),
                       expand = expansion(mult = c(0, 0.05))) +
    labs(x = NULL, y = metric_name) +
    theme_cowplot(font_size = 12) +
    background_grid(major = "y", minor = "none") +
    theme(
      legend.position = "none",
      axis.text.x     = element_text(angle = 60, hjust = 1, vjust = 1),
      strip.background = element_blank(),
      strip.placement  = "outside",
      strip.text       = element_markdown(size = 12),
      panel.spacing    = unit(0.3, "lines"),
      panel.border     = element_rect(colour = "grey70", fill = NA, linewidth = 0.5)
    )
}
```

```{r}
#| fig-width: 12
#| fig-height: 4

walk(metric_order, \(m) print(plot_metric(m)))

```

## Save figures

```{r}

dir.create("./figures", showWarnings = FALSE)

iwalk(figure_stems, \(stem, metric) {
  p <- plot_metric(metric)
  save_plot(glue("./figures/{stem}.pdf"), p, base_width = 12, base_height = 4)
  save_plot(glue("./figures/{stem}.png"), p, base_width = 12, base_height = 4,
            dpi = 300, bg = "white")
})

```

::: {.callout-note collapse="true"}
## Session Info

```{r}
sessionInfo()
```
:::
