---
title: "Collagen qPCR"
subtitle: "Summary plots for paper"
author: "Emily Johnson"
date: today
date-format: short
format:
  html:
    self-contained: true
    theme: litera
    toc: true
editor: source
code-block-bg: true
code-block-border-left: "#31BAE9"
---

# Summary plots

Relative expression of *Col1a1* and *Col1a2* (normalised to *Rps13*) in tendon and
bone across three timepoints, plus the *Col1a1*/*Col1a2* ratio. All six panels
share the same layout, so they are read and plotted through one pipeline.

## Libraries

```{r}
#| output: false
#| warning: false
#| message: false
library(tidyverse)
library(ggbeeswarm)   # beeswarm / quasirandom points
library(cowplot)      # theme_cowplot(), save_plot()
library(ggtext)       # markdown in strips
library(glue)
library(scales)
```

## Read and tidy

One CSV per gene, tissue and timepoint (Windows `latin1`). `Genotype` arrives
quoted (e.g. `"""J +/+"""`); the quotes are stripped and combined with `Sex` into
the x-axis grouping. J = *oim* line, N = *Col1a2*-null.

```{r}
#| warning: false
#| message: false

data_dir   <- "./processed_data/qPCR"
stamp      <- "110726.1"
timepoints <- c("8wks", "18wks", "52wks")

allele_levels <- c(
  "Male J +/+", "Male J +/oim", "Male J oim/oim",
  "Female J +/+", "Female J +/oim", "Female J oim/oim",
  "Male N +/+", "Male N +/-", "Male N -/-",
  "Female N +/+", "Female N +/-", "Female N -/-"
)

expr_files <- tribble(
  ~Gene,    ~Tissue,  ~prefix,
  "Col1a1", "tendon", "Col1a1-tendon",
  "Col1a1", "bone",   "Col1a1-bones",
  "Col1a2", "tendon", "Col1a2-tendon",
  "Col1a2", "bone",   "Col1a2-bones"
)

# `row` preserves file order so Col1a1 and Col1a2 can be paired for the ratio;
# NAs are therefore dropped only after pairing.
read_dataset <- function(prefix) {
  set_names(timepoints) |>
    map(\(tp) {
      read_csv(file.path(data_dir, glue("{prefix}-{tp}_{stamp}.csv")),
               locale = locale(encoding = "latin1"),
               show_col_types = FALSE) |>
        mutate(row = row_number())
    }) |>
    list_rbind(names_to = "Age") |>
    transmute(
      row,
      Sex,
      AlleleLabel = factor(paste(Sex, str_remove_all(Genotype, '"')),
                           levels = allele_levels),
      Timepoint = factor(Age, levels = timepoints,
                         labels = c("8 weeks", "18 weeks", "52 weeks")),
      value = Data
    )
}

expr <- expr_files |>
  mutate(data = map(prefix, read_dataset)) |>
  select(-prefix) |>
  unnest(data)
```

The ratio is derived per sample: the *Col1a1* and *Col1a2* files are matched on
tissue, timepoint and row position (they carry identical genotype/sex in the same
order).

```{r}
ratio <- expr |>
  pivot_wider(names_from = Gene, values_from = value) |>
  transmute(Gene = "Col1a1/Col1a2", Tissue, row, Sex, AlleleLabel, Timepoint,
            value = Col1a1 / Col1a2)

dataset_levels <- c("Col1a1 tendon", "Col1a1 bone",
                    "Col1a2 tendon", "Col1a2 bone",
                    "Col1a1/Col1a2 tendon", "Col1a1/Col1a2 bone")

dat_long <- bind_rows(expr, ratio) |>
  filter(!is.na(value), is.finite(value), value > 0) |>
  mutate(Dataset = factor(paste(Gene, Tissue), levels = dataset_levels))

# Mean ± SD per group
group_stats <- dat_long |>
  summarise(mean = mean(value), sd = sd(value),
            .by = c(Dataset, AlleleLabel, Timepoint))
```

## Plotting

Points are coloured by sex; the sex prefix orders the groups and is stripped from
the tick labels. Expression is plotted on log10, the ratio on log2.

```{r}
sex_cols <- c(Female = "#E69F00", Male = "#0072B2")

figure_stems <- c(
  "Col1a1 tendon"        = "Col1a1-tendon",
  "Col1a1 bone"          = "Col1a1-bone",
  "Col1a2 tendon"        = "Col1a2-tendon",
  "Col1a2 bone"          = "Col1a2-bone",
  "Col1a1/Col1a2 tendon" = "Ratio-tendon",
  "Col1a1/Col1a2 bone"   = "Ratio-bone"
)

plot_dataset <- function(dataset_name, style = c("beeswarm", "jitter")) {

  style <- match.arg(style)
  is_ratio <- str_detect(dataset_name, "/")
  td <- filter(dat_long,    Dataset == dataset_name)
  gd <- filter(group_stats, Dataset == dataset_name)

  points <- switch(style,
    jitter   = geom_jitter(width = 0.1, height = 0, size = 2.5, stroke = 0),
    beeswarm = geom_beeswarm(size = 2.5, cex = 2.2, stroke = 0)
  )

  # mean ± SD: whiskers, wide bar at the mean, narrow bars at ± SD
  whisker <- function(yend) {
    geom_segment(data = gd, aes(x = AlleleLabel, xend = AlleleLabel,
                                y = mean, yend = {{ yend }}),
                 inherit.aes = FALSE, colour = "grey35", linewidth = 0.2)
  }
  crossbar <- function(y, width) {
    geom_crossbar(data = gd, aes(x = AlleleLabel, y = {{ y }},
                                 ymin = {{ y }}, ymax = {{ y }}),
                  inherit.aes = FALSE, width = width,
                  colour = "grey35", linewidth = 0.2)
  }

  y_scale <- if (is_ratio) {
    scale_y_continuous(transform = "log2", limits = c(2^-12, 2^16),
                       breaks = 2^seq(-12, 16, by = 2), labels = label_log(base = 2))
  } else {
    scale_y_log10(limits = c(1e-5, 1e5), breaks = 10^(-5:5), labels = label_log())
  }

  ggplot(td, aes(AlleleLabel, value, colour = Sex)) +
    points +
    whisker(mean + sd) +
    whisker(mean - sd) +
    crossbar(mean, 0.6) +
    crossbar(mean + sd, 0.4) +
    crossbar(mean - sd, 0.4) +
    facet_grid(. ~ Timepoint, scales = "free_x", space = "free_x") +
    scale_colour_manual(values = sex_cols) +
    scale_x_discrete(drop = TRUE, labels = \(x) gsub("^(Male|Female) ", "", x)) +
    y_scale +
    labs(x = NULL,
         y = if (is_ratio) "Relative expression (Col1a2)" else "Relative expression (Rps13)") +
    theme_cowplot(font_size = 12) +
    background_grid(major = "y", minor = "none") +
    theme(
      legend.position  = "none",
      axis.text.x      = element_text(angle = 60, hjust = 1, vjust = 1),
      strip.background = element_blank(),
      strip.placement  = "outside",
      strip.text       = element_markdown(size = 12),
      panel.spacing    = unit(0.3, "lines"),
      panel.border     = element_rect(colour = "grey70", fill = NA, linewidth = 0.5)
    )
}
```

```{r}
#| fig-width: 12
#| fig-height: 4
walk(dataset_levels, \(d) print(plot_dataset(d)))
```

## Save figures

```{r}
dir.create("./figures/qPCR", showWarnings = FALSE)

iwalk(figure_stems, \(stem, dataset_name) {
  p <- plot_dataset(dataset_name)
  save_plot(glue("./figures/qPCR/{stem}.pdf"), p, base_width = 12, base_height = 4)
  save_plot(glue("./figures/qPCR/{stem}.png"), p, base_width = 12, base_height = 4,
            dpi = 300, bg = "white")
})
```

::: {.callout-note collapse="true"}
## Session Info

```{r}
sessionInfo()
```
:::
